---
title: "weather!"
author: "Dylan Morgan"
output: html_document
---

Load packages.
```{r}
library(tidyverse)
library(ggmap)
library(geosphere)
library(lubridate)
library(plotly)
```

Elk data from Wayne.
```{r}
elk = read_csv("./data/elk.csv")
```

all_data csv from Wayne.
```{r}
all_data <- read_csv("./data/all_data.csv")
```

Load and clean up weather data.
```{r}
raw_weather <- read_csv("./data/raw_weather_data.csv")
```

Plot weather stations and elk migration pathways to determine which stations to use.
```{r}
weather_stations <-
  raw_weather |>
  group_by(NAME, LATITUDE, LONGITUDE) |>
  summarize(n_obs = n())

ggplot() +
  geom_path(
    data = elk,
    aes(x=long, y=lat),
    alpha = 0.5,
    color = "red") +
  geom_point(data = weather_stations, aes(x = LONGITUDE, y = LATITUDE))


ggplot(data = weather_stations,
         aes(x = LONGITUDE, y = LATITUDE)) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label = NAME),
                            box.padding   = 0.35,
                            point.padding = 0.5,
                            segment.color = 'grey50',
                            max.overlaps = 20)

potential_stations <- 
  c("LEWIS LAKE DIVIDE, WY US", "SNAKE RIVER, WY US", "SNAKE RIVER STATION, WY US", "BASE CAMP, WY US", 
    "MORAN 5 WNW, WY US", "JACKSON 29.9 NNE, WY US", "BURRO HILL WYOMING, WY US", "MOOSE 1 NNE, WY US", 
    "MOOSE, WY US", "MOOSE .4 S, WY US", "JACKSON 12.3 NE, WY US", "JACKSON 12.2 NE, WY US", 
    "JACKSON HOLE AIRPORT, WY US", "JACKSON, WY US")

four_stations <- 
  c("SNAKE RIVER STATION, WY US", "MORAN 5 WNW, WY US", "BURRO HILL WYOMING, WY US", "MOOSE 1 NNE, WY US")

reduced_weather_stations <- 
  raw_weather |> 
  filter(NAME %in% four_stations) |> 
  group_by(NAME, LATITUDE, LONGITUDE) |> 
  summarize(n_obs = n()) |> 
  arrange(desc(n_obs)) 

four_stations_labels_plot <- 
  ggplot(data = reduced_weather_stations,
         aes(x = LONGITUDE, y = LATITUDE)) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label = NAME),
                            box.padding   = 0.35,
                            point.padding = 0.5,
                            segment.color = 'grey50',
                            max.overlaps = 20)

four_stations_labels_plot

four_stations_elk_mvmt_plot <- 
  ggplot() +
  geom_path(
    data = elk,
    aes(x=long, y=lat),
    alpha = 0.5,
    color = "red") +
  geom_point(data = reduced_weather_stations, aes(x = LONGITUDE, y = LATITUDE))
  
four_stations_elk_mvmt_plot
```

Reduce weather station data, explore weather data.
```{r}
reduced_weather <- 
  raw_weather |> 
  filter(NAME %in% four_stations)
```

What kinds of monthly weather patterns were visible over this time period?

Find patterns in temp, rainfall, snowfall over time. Factor by month. Avg rainfall total, snowfall total, snow depth total, temp avg.

Total precip -- avg totals among the four stations.
Three ways to show data: faceted bar chart per year, point + line graph per year, year-month bar chart separated by year
```{r}
# faceted bar chart, separated by month and year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_precip_total = sum(PRCP, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_precip_total_avg = mean(station_precip_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_precip_total_avg, fill = as.factor(year))) +
  geom_bar(stat = "identity") + 
  facet_wrap(vars(year), ncol = 2) + 
  theme(legend.position = "none") + 
  labs(x = "Month", y = "Total Precipitation (in)")
  
# point + line, separated by month and year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_precip_total = sum(PRCP, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_precip_total_avg = mean(station_precip_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_precip_total_avg, color = year, group = year)) + 
  geom_point(size = 2) + 
  geom_line(linewidth = 1) + 
  labs(x = "Month", y = "Total Precipitation (in)", color = "Year")
  
# year-month totals, separated by year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_precip_total = sum(PRCP, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_precip_total_avg = mean(station_precip_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = year_month, y = station_precip_total_avg, fill = as.factor(year))) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Date (Year, Month)", y = "Total Precipitation (in)", fill = "Year")
  
```

Precip plotly.
```{r}
# point + line, separated by month and year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_precip_total = sum(PRCP, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_precip_total_avg = mean(station_precip_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~month, y = ~station_precip_total_avg, color = ~year, group = ~year, type = 'scatter', mode = 'lines'
    ) |> 
  layout(xaxis = list(title = "Month"), 
         yaxis = list(title = "Total Precipitation (in)"), legend = list(title = list(text = "Year")))

# year-month totals, separated by year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_precip_total = sum(PRCP, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_precip_total_avg = mean(station_precip_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~year_month, y = ~station_precip_total_avg, color = ~as.factor(year), type = "bar"
    ) |> 
  layout(xaxis = list(title = "Date (Year, Month)"), 
         yaxis = list(title = "Total Precipitation (in)"), legend = list(title = list(text = "Year")))
```

Snowfall.
```{r}
# faceted bar chart, separated by month and year -- KEEP STATIC, USE FOR ANALYSIS PAGE
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowfall_total = sum(SNOW, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowfall_total_avg = mean(station_snowfall_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_snowfall_total_avg, fill = as.factor(year))) +
  geom_bar(stat = "identity") + 
  facet_wrap(vars(year), ncol = 2) + 
  theme(legend.position = "none") + 
  labs(x = "Month", y = "Total Snowfall (in)")
  
# point + line -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowfall_total = sum(SNOW, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowfall_total_avg = mean(station_snowfall_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_snowfall_total_avg, color = year, group = year)) + 
  geom_point(size = 2) + 
  geom_line(linewidth = 1) + 
  labs(x = "Month", y = "Total Snowfall (in)", color = "Year")

# year-month totals by year -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowfall_total = sum(SNOW, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowfall_total_avg = mean(station_snowfall_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = year_month, y = station_snowfall_total_avg, fill = as.factor(year))) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Date (Year, Month)", y = "Total Snowfall (in)", fill = "Year")
```

Snowfall plotly.
```{r}
# point + line
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowfall_total = sum(SNOW, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowfall_total_avg = mean(station_snowfall_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~month, y = ~station_snowfall_total_avg, color = ~as.factor(year), group = ~year, type = 'scatter', mode = 'lines'
    ) |> 
  layout(xaxis = list(title = "Month"), 
         yaxis = list(title = "Total Snowfall (in)"), legend = list(title = list(text = "Year")))

# year-month totals by year
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowfall_total = sum(SNOW, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowfall_total_avg = mean(station_snowfall_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~year_month, y = ~station_snowfall_total_avg, color = ~year, type = 'bar'
    ) |> 
  layout(xaxis = list(title = "Date (Year, Month)"), 
         yaxis = list(title = "Total Snowfall (in)"), legend = list(title = list(text = "Year")))
```

Snow depth.
```{r}
# faceted bar chart, separated by month and year -- KEEP STATIC, USE FOR ANALYSIS PAGE
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowdepth_total = sum(SNWD, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowdepth_total_avg = mean(station_snowdepth_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_snowdepth_total_avg, fill = as.factor(year))) +
  geom_bar(stat = "identity") + 
  facet_wrap(vars(year), ncol = 2) + 
  theme(legend.position = "none") + 
  labs(x = "Month", y = "Total Snow Depth (in)")

# point + line -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowdepth_total = sum(SNWD, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowdepth_total_avg = mean(station_snowdepth_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = station_snowdepth_total_avg, color = year, group = year)) + 
  geom_point(size = 2) + 
  geom_line(linewidth = 1) + 
  labs(x = "Month", y = "Total Snow Depth (in)", color = "Year")

# year-month totals -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowdepth_total = sum(SNWD, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowdepth_total_avg = mean(station_snowdepth_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = year_month, y = station_snowdepth_total_avg, fill = as.factor(year))) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Date (Year, Month)", y = "Total Snow Depth (in)", fill = "Year")
```

Snow depth plotly.
```{r}
# point + line
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowdepth_total = sum(SNWD, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowdepth_total_avg = mean(station_snowdepth_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~month, y = ~station_snowdepth_total_avg, color = ~as.factor(year), group = ~year, type = 'scatter', mode = 'lines'
    ) |> 
  layout(xaxis = list(title = "Month"), 
         yaxis = list(title = "Total Snow Depth (in)"), legend = list(title = list(text = "Year")))

# year-month totals
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(NAME, year, month) |> 
  summarize(station_snowdepth_total = sum(SNWD, na.rm = TRUE)) |> 
  group_by(year, month) |> 
  summarize(station_snowdepth_total_avg = mean(station_snowdepth_total, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~year_month, y = ~station_snowdepth_total_avg, color = ~year, type = 'bar'
  ) |> 
  layout(xaxis = list(title = "Date (Year, Month)"), 
         yaxis = list(title = "Total Snow Depth (in)"), legend = list(title = list(text = "Year")))
```

Avg temp. 
```{r}
# faceted bar chart, separated by month and year -- KEEP STATIC, USE FOR ANALYSIS PAGE
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(year, month) |> 
  summarize(year_month_tavg = mean(TAVG, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = year_month_tavg, fill = as.factor(year))) +
  geom_bar(stat = "identity") + 
  facet_wrap(vars(year), ncol = 2) + 
  theme(legend.position = "none") + 
  labs(x = "Month", y = "Average Temperature (F)")

# point + line -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(year, month) |> 
  summarize(year_month_tavg = mean(TAVG, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = month, y = year_month_tavg, color = year, group = year)) + 
  geom_point(size = 2) + 
  geom_line(linewidth = 1) + 
  labs(x = "Month", y = "Average Temperature (F)", color = "Year")

# year-month totals -- MAKE DYNAMIC, USE FOR DASHBOARD
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(year, month) |> 
  summarize(year_month_tavg = mean(TAVG, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  ggplot(aes(x = year_month, y = year_month_tavg, fill = as.factor(year))) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Date (Year, Month)", y = "Average Temperature (F)", fill = "Year")
```

Avg temp plotly.
```{r}
# point + line
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(year, month) |> 
  summarize(year_month_tavg = mean(TAVG, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~month, y = ~year_month_tavg, color = ~as.factor(year), group = ~year, type = 'scatter', mode = 'lines'
    ) |> 
  layout(xaxis = list(title = "Month"), 
         yaxis = list(title = "Average Temperature (F)"), legend = list(title = list(text = "Year")))

# year-month totals
reduced_weather |> 
  select(c(STATION, NAME, LATITUDE, LONGITUDE, DATE, PRCP, SNOW, SNWD, TAVG)) |> 
  mutate(year = format(DATE, "%Y"), 
         month = format(DATE, "%m")) |> 
  group_by(year, month) |> 
  summarize(year_month_tavg = mean(TAVG, na.rm = TRUE)) |> 
  mutate(year_month = paste(year, month, sep = "-")) |> 
  plot_ly(
    x = ~year_month, y = ~year_month_tavg, color = ~year, type = 'bar'
  ) |> 
  layout(xaxis = list(title = "Date (Year, Month)"), 
         yaxis = list(title = "Average Temperature (F)"), legend = list(title = list(text = "Year")))
```

Scatterplot of daily elk movement versus precip
Scatterplot of daily elk movement versus snowfall
Scatterplot of daily elk movement versus snow depth
Scatterplot of daily elk movement versus temperature

Daily elk.
```{r}
daily_elk <- 
  all_data |> 
  mutate(year_month_day = format(dt, "%Y-%m-%d"), 
         elk_id = as.character(elk_id)) |> 
  group_by(
    elk_id,
    year_month_day, 
    year,
    month,
    day,
    tavg,
    prcp,
    snow,
    snwd,
    tmin,
    tmax
  ) |>
  summarize(
    dist_km = sum(dist_km, na.rm = TRUE), 
    land_cover = mean(land_cover, na.rm = TRUE)
  )
```

*FOR ALL OF THE PLOTS BELOW, MAKE STATIC AND DYNAMIC, AND USE ON BOTH ANALYSIS PAGE AND DASHBOARD*

Elk mvmt vs precip.
```{r}
elk_prcp_distkm_pts <- 
  daily_elk |> 
  ggplot(aes(x = prcp, y = dist_km)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Precipitation (in)", y = "Daily Distance Traveled (km)")

style(elk_prcp_distkm_pts, hoverinfo = "none", traces = 2)

elk_prcp_distkm_smooth <- 
  daily_elk |> 
  ggplot(aes(x = prcp, y = dist_km, color = elk_id)) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Precipitation (in)", y = "Daily Distance Traveled (km)", color = "Elk ID")

ggplotly(elk_prcp_distkm_smooth)
```

Elk mvmt vs. snowfall.
```{r}
elk_snowfall_distkm_pts <- 
  daily_elk |> 
  ggplot(aes(x = snow, y = dist_km)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Snowfall (in)", y = "Daily Distance Traveled (km)")

style(elk_snowfall_distkm_pts, hoverinfo = "none", traces = 2)

elk_snowfall_distkm_smooth <- 
  daily_elk |> 
  ggplot(aes(x = snow, y = dist_km, color = elk_id)) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Snowfall (in)", y = "Daily Distance Traveled (km)", color = "Elk ID")

ggplotly(elk_snowfall_distkm_smooth)
```

Elk mvmt vs. snow depth. 
```{r}
elk_snowdepth_distkm_pts <- 
  daily_elk |> 
  ggplot(aes(x = snwd, y = dist_km)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Snow Depth (in)", y = "Daily Distance Traveled (km)")

style(elk_snowdepth_distkm_pts, hoverinfo = "none", traces = 2)

elk_snowdepth_distkm_smooth <- 
  daily_elk |> 
  ggplot(aes(x = snwd, y = dist_km, color = elk_id)) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Daily Snow Depth (in)", y = "Daily Distance Traveled (km)", color = "Elk ID")

ggplotly(elk_snowdepth_distkm_smooth)
```

Elk mvmt vs avg temp.
```{r}
elk_avgtemp_distkm_pts <- 
  daily_elk |> 
  ggplot(aes(x = tavg, y = dist_km)) + 
  geom_point() + 
  geom_smooth(se = FALSE) + 
  labs(x = "Average Daily Temperature (F)", y = "Daily Distance Traveled (km)")

style(elk_avgtemp_distkm_pts, hoverinfo = "none", traces = 2)

elk_avgtemp_distkm_smooth <- 
  daily_elk |> 
  ggplot(aes(x = tavg, y = dist_km, color = elk_id)) + 
  geom_smooth(se = FALSE) + 
  labs(x = "Average Daily Temperature (F)", y = "Daily Distance Traveled (km)", color = "Elk ID")

ggplotly(elk_avgtemp_distkm_smooth)
```




