---
title: "Analysis Starter Code"
author: "Wayne Monical"
date: "2024-12-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

library load
```{r}
library(tidyverse)
```



Read in main data set. Remember to set your working directory to the folder with this Rmd file! You can do this by navigating to your folder in the Rstudio `Files` tab, clicking the `gear` icon, and clicking the `Set as working directory` button.
```{r}
all_data = read_csv('../clean_data/all_data.csv')
```


### Analyzing elk movement 

Analyze elk movement with the `dist_km` column. Remember to drop the NA's!
```{r}
all_data |> 
  drop_na(dist_km) |> 
  filter(dist_km < 10) |> 
  ggplot(aes(x = dist_km))+
  geom_histogram()
```



### Analyzing all water quality data

You can analyze the orignal water quality data in a wide or a long format
```{r}
# wide format
clean_water = read_csv('../clean_data/clean_water.csv')

# long format
water_quality = read_csv('../clean_data/water_quality.csv')
```


