---
title: "Analysis"
author: "Wayne Monical"
date: "2024-11-17"
output: html_document
---


## Load Packages
```{r}
library(tidyverse)
library(ggmap)
```



## Read in Data

```{r}
elk = read_csv('../clean_data/elk.csv')
water_quality = read_csv('../clean_data/water_quality.csv')
```

## Elk Data

Each elk has an `elk_id`. The `year`, `month`, `day`, and `hour` give the time. The `lat` and `long` give the location.

```{r}
head(elk)
```
## Water Quality

- The `year`, `month`, and `day` variables give the time. 
- The `latitude` and `longitude` give the location. 
- `characteristic_name` gives the type of measurement. 
- `result_text` gives the measurement. 

```{r}
head(water_quality)
```

This is an incredibly rich data set. I have only kept the 20 most common quantitative measurements, but there are so many more. I have kept this data set in the long format, because there is very spotty measuring. 

```{r}
water_quality |> 
  group_by(characteristic_name) |> 
  summarize(n = n()) |> 
  arrange(desc(n))
```


### Water Quality pH

This code chunk filters for all measurements of pH. `result_text` is a character vector, so make sure to change the vector to a numeric type if you're working with numeric data, like pH

```{r}
calcium = 
  water_quality |> 
  filter(characteristic_name == 'Calcium mg/l') |> 
  mutate(calcium = as.numeric(result_text)) # turning the vector numeric


calcium |>    
  ggplot(aes(x = calcium)) + 
  geom_histogram()
```


## Create a static map plot

[using ggmap](https://www.nceas.ucsb.edu/sites/default/files/2020-04/ggmapCheatsheet.pdf)

[from stack exchange](https://stackoverflow.com/questions/33942186/using-r-and-ggplot-to-draw-separate-lines-between-gps-coordinates/33944974)


Filter for a single elk named alice. We'll use her for testing.
```{r}
alice = 
  elk |> 
  filter(
    elk_id == 572,
    year %in% c(2006, 2007)
  )
```

### Step 1: Download map

Find the minimum and maximum latitude and longitude of Bob's journey. This will give us the range of map to download.

```{r}
min_lat = alice |> pull(lat) |> min()
max_lat = alice |> pull(lat) |> max()
rng_lat = abs(min_lat - max_lat)
lowerleftlat = min_lat 
upperrightlat = max_lat 


min_long = alice |> pull(long) |> min()
max_long = alice |> pull(long) |> max()
rng_long = abs(min_long - max_long)
lowerleftlon = min_long - rng_long 
upperrightlon = max_long + rng_long

myLocation <- c(left = lowerleftlon,
                 bottom = lowerleftlat,
                 right = upperrightlon,
                 top = upperrightlat)
```


```{r}
register_stadiamaps(key = '29074900-bb6e-4a71-8f91-454c28190f88', write = FALSE)

myMap <- get_stadiamap(
  bbox=myLocation,
  maptype = "stamen_terrain",
  crop=FALSE)
```

### Step 2: Plot map with elk movement
```{r}
ggmap(myMap) +
geom_path(
  data = alice, 
  aes(x=long, y=lat, color = month))+
  geom_line(alpha = 0) +
  scale_color_gradientn(colours = rainbow(12))
```


```{r}
ggplot(
  data = elk, 
  aes(x=long, y=lat)) +
geom_path(alpha = 0.5) +
geom_line(alpha = 0) +
geom_point(
  data = calcium,
  aes(x = longitude, y = latitude, color = 'red')
  )
```

Add in water calcium
```{r}
ggmap(myMap) +
geom_path(
  data = elk, 
  aes(x=long, y=lat, color = month))+
  geom_line(alpha = 0) +
  scale_color_gradientn(colours = rainbow(12)) +
geom_point(
  data = calcium,
  aes(x = longitude, y = latitude)
)
```



